Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-10-14 paper/submitted/round3_png/algorithms_v6_cmame.tex
more edits

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena (as in the example above), or be a reduced-order model, or have a reduced parameter and/or state space. The derived error estimate is not applicable to all models, however. The low- and high-fidelity models must be expressible in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. In addition, as described in \Cref{sec:setup}, the models must be compatible, in that the low-fidelity parameters and states $q_{LF}$ and $u_{LF}$ must lie in the high-fidelity parameter and state spaces as well. This implies the two models must share the same domain; for example, the low-fidelity model can not be a one-dimensional average of a two-dimensional high-fidelity model. Models that differ in discretization can be compatible; the low-fidelity model may have a coarser mesh, basis functions of a lower order, or basis functions spanning a reduced subspace. Models that have different states and/or parameters can also be compatible; for example, the high-fidelity model could describe changes in states that have a fixed value (often zero) in the low-fidelity model. The high-fidelity model could allow anisotropy in its parameters while the low-fidelity model could constrain their components and require isotropy. We note that in the case where the two models correspond to two levels of mesh refinement, method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead.

>>>>

\Cref{alg:refSeries} is applicable to many, though not all, models. One restriction to note is that the low- and high-fidelity models must be expressible in a weak form, so \Cref{alg:refSeries} cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. The models must also be compatible (see \Cref{sec:setup}). Note that this implies the two models must share the same domain; for example, the low-fidelity model can not be a one-dimensional average of a two-dimensional high-fidelity model.

The requirements for compatibility are loose enough so that \Cref{alg:refSeries} is applicable to a large class of models. Models that differ in discretization can be compatible; the low-fidelity model may have a coarser mesh, basis functions of a lower order, or basis functions spanning a reduced subspace. Models that have different states and/or parameters can also be compatible; for example, the high-fidelity model could describe changes in states that have a fixed value (often zero) in the low-fidelity model. The high-fidelity model could allow anisotropy in its parameters while the low-fidelity model could constrain their components and require isotropy. We note that in the case where the two models correspond to two levels of mesh refinement, method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead.

### 2017-10-14 paper/submitted/round3_png/mathematical_formulation_v6_cmame.tex
more edits

A given physical system can be described to varying degrees of fidelity using different models. One has a choice in how to define the structure of the inferred parameters $q$, as well as how to define the state equation and variables. For example, the parameter $q$ can be modeled as a single scalar value or as a distributed field variable. The state equation can be nonlinear and multiscale, requiring an expensive computation to resolve (e.g.\ the Navier-Stokes equations) or a linear, single physics model (e.g.\ Laplace's equation). All these different modeling choices lead to different versions of the optimization problem described by \Cref{eq:invOpt} that would need to be solved to infer the unknown parameters. Note that in order to solve \Cref{eq:invOpt} numerically, the description of a model must usually include a choice of discretization method in addition to a choice of equations.

@@

Assume there is compatibility between the low-fidelity and high-fidelity parameter and state solution spaces; specifically, assume that the low-fidelity parameters and states $q_{LF}$ and $u_{LF}$ lie in the high-fidelity parameter and state spaces as well, so that we have $q_{LF}\in Q_{HF}$ and $u_{LF}\in U_{HF}$ (see \Cref{sec:adapt_alg} for interpretation details). Given this compatibility, one may combine the two models to form a mixed-fidelity model and corresponding mixed-fidelity inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of compatible models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution; we now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the forms $a_{HF/LF}:U_{HF/LF} \times Q_{HF/LF} \times U_{HF/LF} \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the QoI functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

>>>>

A given physical system can be described to varying degrees of fidelity using different models. One has a choice in how to define the structure of the inferred parameters $q$, as well as how to define the state equation and variables. For example, the parameter $q$ can be modeled as a single scalar value or as a distributed field variable. The state equation can be nonlinear and multiscale, requiring an expensive computation to resolve (e.g.\ the Navier-Stokes equations) or a linear, single physics model (e.g.\ Laplace's equation). All these different modeling choices lead to different versions of the optimization problem described by \Cref{eq:invOpt} that would need to be solved to infer the unknown parameters. In practice, \Cref{eq:invOpt} often has no analytical solutions and must be solved numerically, in which case the description of a model includes a choice of discretization method in addition to a choice of equations.

@@

Assume the high- and low-fidelity models are \emph{compatible}, in that $U_{LF}\subseteq U_{HF}$ and $Q_{LF}\subseteq Q_{HF}$. Then this compatibility allows one to combine the two models to form a mixed-fidelity model and corresponding mixed-fidelity inverse problem. The low-fidelity model can be used in part of the computational domain, and the high fidelity in the remainder; we denote the mixed-fidelity model and its variables with the subscript $_{MF}$. How one chooses to combine the low- and high-fidelity models can be informed by what aspect of the inferred parameters is of interest.

@@

%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of compatible models, we take the QoI calculated from inferring the parameters with the highest-fidelity model to be the `exact' or `true' solution; we now derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{proposition}
\label{thm:error_estimate}
Consider the high- and low-fidelity inverse problems described by the constrained optimization problems in \Cref{eq:invOptHF,eq:invOptLF}. Let the high- and low-fidelity models be compatible ($U_{LF}\subseteq U_{HF}$ and $Q_{LF}\subseteq Q_{HF}$), and let the forms $a_{HF/LF}:U_{HF/LF} \times Q_{HF/LF} \times U_{HF/LF} \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U_{HF/LF}\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q_{HF/LF}\to\Reals$ be differentiable with respect to the parameter $q$, and the QoI functional $I:Q_{HF/LF}\times U_{HF/LF}\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.
